---
title: "Reproducible Research: Course Project #1"
author: "Josh Choate"
date: "March 14, 2018"
output: html_document
keep_md: true
---
The following is a brief analysis of activity monitoring data. This is my attempt at Course Project #1 in the Coursera Reproducible Research on-line course by Johns Hopkins University.

First, the necessary R packages and the activity monitoring data are loaded:

```{r warning=FALSE}
library(dplyr)
library(ggplot2)
library(reshape2)

activity_data <- read.csv("C:/Users/JoCho/Desktop/activity.csv")

str(activity_data)

activity_data$date <- as.Date(activity_data$date)
```

Next, the assigned tasks (numbered 1 - 8) are performed.

1. Histogram of the total number of steps taken each day:
```{r}
totals <- activity_data %>%
          group_by(date) %>%
          summarise(steps = sum(steps, na.rm = TRUE))
 
ggplot(totals, aes(steps)) + geom_histogram(fill = "steelblue", color = "black", bins = 40) +
      labs(title = "Total Steps Taken", y = "days")
```


2. Mean and Median of the total number of steps taken per day:
```{r}
head(totals) %>%
   summarise(mean = mean(steps, na.rm=TRUE), median = median(steps, na.rm=TRUE))
    
```


3. Time series plot of the average number of steps taken:
```{r}
totals$average <- totals$steps/length(totals$date)

plot(totals$date, totals$average, main = "Daily Step Count Averages", xlab = "Date", 
     ylab = "Average Number of Steps per Five Minutes")


```


4. The 5-minute interval that, on average, contains the maximum number of steps:
```{r}
intervals <- activity_data %>% group_by(interval) %>%
  summarise(mean = mean(steps, na.rm=TRUE))

max_int <- max(intervals$mean)

intervals[intervals$mean == max_int,]

```


5. Code to describe and show a strategy for imputing missing data:

First, how much missing data is there?
```{r}
summary(activity_data$steps)

NA_ratio = sum(is.na(activity_data$steps))/length(activity_data$steps)
NA_ratio
```

13% is a significant figure, but amounts to less than 5 of the 61 days of data. Are these 5+ days of data reflective of certain days when the tracker was not worn, or are the missing values randomly scattered? 
```{r}
ggplot(activity_data, aes(x=date, y=steps)) + 
  geom_line() +
  scale_x_date(date_breaks = '10 days', date_labels = '%m/%d')

```

The above plot shows that the large majority of NA values occurred on specific days. Notably, there is no data for Oct.1 and a small amout for October 2. Thus, it appears likely that the subject did not begin tracking his/her steps until late on Oct. 2. Consequently, those two dates will be omitted from the set. The remaining values will be imputed using the mice package. (Against the assignment protocal, I've set results="hide" here to avoid showing the mice function's lengthy output.)

```{r warning=FALSE, results="hide"}
library(mice)

activity_data_NArm <- filter(activity_data, date > "2012-10-02")
steps_only <- data.frame(activity_data_NArm[,1], activity_data_NArm[,3])

temp <- mice(data = steps_only, m = 5, method = "pmm", maxit = 50, seed = 500)

complete_steps <- complete(temp,1)

activity_data_NArm$steps <- complete_steps[,1]
  
```


7. Histogram of total number of steps per day once missing values are replaced:
```{r}
totals_NArm <- activity_data_NArm %>%
               group_by(date) %>%
               summarise(steps = sum(steps, na.rm = TRUE))
 
ggplot(totals_NArm, aes(steps)) + geom_histogram(fill = "green", color = "black", bins = 40) +
      labs(title = "Total Steps Taken (NA Values Removed)", y = "days")
```


8. Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends:
```{r}
totals$weekend <- grepl("S.+",weekdays(totals$date))

weekday <- filter(totals, weekend == FALSE)
weekend <- filter(totals, weekend == TRUE)

weekday$average <- weekday$steps/length(weekday$date)
weekend$average <- weekend$steps/length(weekend$date)

par(mfrow = c(1, 2), mar = c(5, 4, 2, 1))
with(weekday, plot(date, steps, type = "l", main = "Weekday Steps"))
with(weekend, plot(date, steps, type = "l", main = "Weekend Steps"))
```


Thank you and good night.
